---
title: "MLM Nested Main Section B"
author: "Xinming Dai; Checked by Chongjun Liao, Jeremy Lu, Yu Wang "
header-includes:
   - \usepackage{todonotes}
output:
  pdf_document: default
date: "Compiled on `r format(Sys.time(), '%a %b %d %H:%M %Z')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,tidy=TRUE)
require(lme4)
require(foreign)
require(lmerTest)
require(tidyverse)
```

## Question 0: read data and process missingness

```{r question 0}
dat <- read_csv("classroom.csv")

# construct new outcome math1st
dat <- 
  dat %>% 
  mutate(math1st = mathkind + mathgain)

# remove missing data
dat <- 
  dat %>% 
  filter(complete.cases(dat))
```

## Question 1
```{r question 1}
# fit a model
fit1 <- lmerTest::lmer(math1st ~ housepov + yearstea + mathprep + mathknow + ses + sex + minority + (1|schoolid/classid), data = dat)
summary(fit1)

# plot residuals to test normality assumption
res1 <- residuals(fit1)
qqnorm(res1, pch = 1, frame = FALSE)
qqline(res1, col = "steelblue", lwd = 2)
```

QQ plot shows that points are around the line, and thus we believe the normality assumption holds.

## Question 2

```{r question 2}
# Generate the two sets of BLUPs (for random effects zeta0 and eta0)
blups_fit1 <- ranef(fit1)

par(mfrow=c(1,2))  
# examine normality for eta0 (class-level)
eta0_fit1 <- blups_fit1$`classid:schoolid`$`(Intercept)`
qqnorm(eta0_fit1, pch = 1, frame = FALSE, main = "Normal Q-Q plot for eta0")
qqline(eta0_fit1, col = "steelblue", lwd = 2)
  
# examine normality for zeta0 (school-level)
zeta0_fit1 <- blups_fit1$schoolid$`(Intercept)`
qqnorm(zeta0_fit1, pch = 1, frame = FALSE, main = "Normal Q-Q plot for zeta0")
qqline(zeta0_fit1, col = "red", lwd = 2)
par(mfrow=c(1,1))  
```

QQ plot shows that both sets of BLUPs of zeta0 and eta0 are around the line, and thus we believe the normality assumption holds.

## Question 3

```{r question 3}
# a
# add a random slope for minority, correlated with the random intercept, at the school level
fit2 <- lmerTest::lmer(math1st ~ housepov + yearstea + mathprep + mathknow + ses + sex + minority + (minority|schoolid) + (1 | classid), data = dat)
print(summary(fit2))

# b
# residual
blups_fit2 <- ranef(fit2)
# BULPs
zeta0 <- blups_fit2$schoolid$`(Intercept)`
zeta1 <- blups_fit2$schoolid$minority
eta0 <- blups_fit2$classid$`(Intercept)`

# c
# check normality
par(mfrow=c(2,2))  
# examine normality for eta0 (class-level)
qqnorm(eta0, pch = 1, frame = FALSE, main = "Normal Q-Q plot for eta0")
qqline(eta0, col = "steelblue", lwd = 2)
  
# examine normality for zeta0 (school-level)
zeta0_fit1 <- blups_fit1$schoolid$`(Intercept)`
qqnorm(zeta0, pch = 1, frame = FALSE, main = "Normal Q-Q plot for zeta0")
qqline(zeta0, col = "red", lwd = 2)

# examine normality for zeta1 (random slop)
qqnorm(zeta1, pch = 1, frame = FALSE, main = "Normal Q-Q plot for zeta1")
qqline(zeta1, col = "chartreuse3", lwd = 2)
par(mfrow=c(1,1)) 
```
QQ plot shows that BLUPs of zeta0, eta0, and zeta1 are around the line, and thus we believe the normality assumption holds.

```{r}
# d
plot(zeta1, zeta0)
```
Overall, zeta0 and zeta1 are negative correlated. However, some odd points are positive correlated.

```{r}
# e
# these schools are 
unique(dat$schoolid)[3-abs(blups_fit2$schoolid$minority)>0]
odd_point <- 
  dat %>% 
  filter(schoolid %in% unique(dat$schoolid)[3-abs(blups_fit2$schoolid$minority)>0])
```
Students in these schools are almost minority.

## Question 4
### a
$V_S = 93.89$, $V_C = 169.45$, and $V_E = 1064.96$.

```{r Question 4}
# fit a model
fit3 <- lmerTest::lmer(math1st ~ housepov + yearstea + mathprep + mathknow + ses + sex + minority + (ses|schoolid) + (1 | classid), data = dat)
summary(fit3)
```
### c
$V_C=86.57$, $V_{S(ses=0)}=171.18$, and $V_E=73.36$.

### d
$V_{S(ses=-0.50)}=171.18+2*(-0.5)*13.083*8.565*0.19+(-0.5)^2*73.36=$ `r round(171.18+2*(-0.5)*13.083*8.565*0.19+(-0.5)^2*73.36, 2)`

$V_{S(ses=0.50)}=171.18+2*(0.5)*13.083*8.565*0.19+(0.5)^2*73.36=$ `r round(171.18+2*(0.5)*13.083*8.565*0.19+(0.5)^2*73.36, 2)`

### e
There is heteroscedasticity at school level because $V_{S(ses=0.50)}$ and $V_{S(ses=-0.50)}$ are not approximate.

