---
title: "MLM Nested Project D"
author: "Xinming Dai, Chongjun Liao, Jeremy Lu, Yu Wang "
header-includes:
   - \usepackage{todonotes}
output:
  pdf_document: default
date: "Compiled on `r format(Sys.time(), '%a %b %d %H:%M %Z')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,tidy=TRUE)
require(lme4)
require(foreign)
require(lmerTest)
require(tidyverse)
```

## Question 1: data generating process

```{r question 1}
set.seed(2042001)

# variance of the random effect
sigma_eta_2 <- 2
sigma_epsilon_2 <- 2

# generate data
dat <- 
  tibble(classid = rep(c(1:100), each = 200),
         studentid = 1:(100*200),
         x = runif(100*200, min = 0, max = 1),
         eta_j = rep(rnorm(100, sd = sqrt(sigma_eta_2)), each = 200),
         epsilon = rnorm(100*200, sd = sqrt(sigma_epsilon_2)),
         y = x + eta_j + epsilon)

```

# Question 2: fit the model

```{r fit1}
lmer_fit1 <- lmer(y ~ x + (1|classid), data = dat)
summary(lmer_fit1)
```

## Question 2: 
  a. The estimated coefficient of X is 0.986.
  b. The 95% confidence interval for this coefficient estimate is $[0.986-1.96*0.035, 0.986+1.96*0.035] =$ $[$ `r 0.986-1.96*0.035`$,$ `r 0.986+1.96*0.035`$]$. It covers the true coefficient, which is 1.

## Question 3: 

## Question 4: 
```{r}
# 4a
z <- rbinom(100*200,1,dat$x)
# 4b
dat$y_q4 <- ifelse(z==1,NA,dat$y)
# 4c
lmer_fit_mar <- lmer(y_q4 ~ x + (1|classid), data = dat)
summary(lmer_fit_mar)
```

### d.
  i. The 95% confidence interval is [`r round(as.data.frame(coef(summary(lmer_fit_mar)))[2,1]-1.96*as.data.frame(coef(summary(lmer_fit_mar)))[2,2],3)`,`r round(as.data.frame(coef(summary(lmer_fit_mar)))[2,1]+1.96*as.data.frame(coef(summary(lmer_fit_mar)))[2,2],3)`], which covers the "truth".  
  
### e.
```{r}
N <- sum(z==0)
```
We use N = `r N` samples in the model fit.


## Question 5: 




