---
title: "MLM Nested Project D"
author: "Xinming Dai, Chongjun Liao, Jeremy Lu, Yu Wang "
header-includes:
  - \usepackage{todonotes}
output:
  pdf_document: default
date: "Compiled on `r format(Sys.time(), '%a %b %d %H:%M %Z')`"
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,tidy=TRUE)
require(lme4)
require(foreign)
require(lmerTest)
require(tidyverse)
```

## Question 1: data generating process

```{r question 1}
set.seed(2042001)
# variance of the random effect
sigma_eta_2 <- 2
sigma_epsilon_2 <- 2
# generate data
dat <- 
  tibble(classid = rep(c(1:100), each = 200),
         studentid = 1:(100*200),
         x = runif(100*200, min = 0, max = 1),
         eta_j = rep(rnorm(100, sd = sqrt(sigma_eta_2)), each = 200),
         epsilon = rnorm(100*200, sd = sqrt(sigma_epsilon_2)),
         y = x + eta_j + epsilon)
```

# Question 2: fit the model

```{r fit1}
lmer_fit1 <- lmer(y ~ x + (1|classid), data = dat)
summary_lmer_fit1 <- summary(lmer_fit1)
summary_lmer_fit1
estimate_x <- summary_lmer_fit1$coefficients[2, 1]
se_x <- summary_lmer_fit1$coefficients[2, 2]
```

## Question 2: 
a. The estimated coefficient of X is `r round(estimate_x, 4)`.
b. The 95% confidence interval for this coefficient estimate is $[0.986-1.96*0.035, 0.986+1.96*0.035] = [`r round(estimate_x-1.96*se_x, 4)`, `r round(estimate_x+1.96*se_x, 4)`]$. It covers the true coefficient, which is 1.

## Question 3: 
```{r}
# 3a
dat_copy <- dat
# 3b
Z_Q3 <- rbinom(20000, 1, 0.5)
table(Z_Q3)
# 3c
dat_copy <- dat_copy %>% mutate(y = replace(y, 1:n(), ifelse(Z_Q3==1, NA, y)))
# 3d
lmer_fit_Q3 <- lmer(y ~ x +(1|classid), data = dat_copy)
summary(lmer_fit_Q3)
# 3f
N_Q3 <- nrow(dat)-sum(is.na(dat_copy$y))
N_Q3
```

### e.
The estimate coefficient is 1.02, which does not change too much
The 95% CI is $[1.02-1.96*0.05, 1.02+1.96*0.05]$, which is $[0.92, 1.12]$ almost converges to the true value

### f.
The total sample size used in this Question is 9945


## Question 4: 
```{r}
# 4a
dat_copy_4 <- dat
z <- rbinom(100*200,1,dat_copy_4$x)
table(z)
# 4b
dat_copy_4$y <- ifelse(z==1,NA,dat_copy_4$y)
# 4c
lmer_fit_4 <- lmer(y ~ x + (1|classid), data = dat_copy_4)
summary_lmer_fit4 <- summary(lmer_fit_4)
summary_lmer_fit4

estimate_x <- summary_lmer_fit4$coefficients[2, 1]
round(estimate_x, 4)
```

### d.
i. The 95% confidence interval is [`r round(as.data.frame(coef(summary(lmer_fit_4)))[2,1]-1.96*as.data.frame(coef(summary(lmer_fit_4)))[2,2],3)`,`r round(as.data.frame(coef(summary(lmer_fit_4)))[2,1]+1.96*as.data.frame(coef(summary(lmer_fit_4)))[2,2],3)`], which covers the "truth".  

### e.
```{r}
N <- nrow(dat)-sum(is.na(dat_copy_4$y))
```
We use N = `r N` samples in the model fit.


## Question 5: 
```{r}
dat_copy_5 <- dat
### a
expit <- function(x){exp(x)/(1+exp(x))}

### b
z <- rbinom(100*200,1, expit(dat_copy_5$y))
table(z)
### c
dat_copy_5$y <- ifelse(z==1,NA,dat_copy_5$y)

### d
lmer_fit_5 <- lmer(y ~ x + (1|classid), data = dat_copy_5)
summary(lmer_fit_5)
```
The new estimate for slope is `r round(as.data.frame(coef(summary(lmer_fit_5)))[2,1], 3)`.

### e
The 95% confidence interval is [`r round(as.data.frame(coef(summary(lmer_fit_5)))[2,1]-1.96*as.data.frame(coef(summary(lmer_fit_5)))[2,2],3)`,`r round(as.data.frame(coef(summary(lmer_fit_5)))[2,1]+1.96*as.data.frame(coef(summary(lmer_fit_5)))[2,2],3)`], which does not cover the "truth", besides the intercept also change.  

### f
The total sample size is `r nrow(dat_copy_5)-sum(is.na(dat_copy_5$y))`, based on number of observations.
